Potential competition in alpine bumble bees is mitigated by imperfect trait-matching and subpartitioning within floral morphotypes

### 2021-03-26 ms_1/ms_1.Rmd
almost done ms2...

Bumble bees (Hymenoptera: *Bombus* spp.) are classic model organisms in the field of community ecology, but they have more recently become conservation priorities due to well-documented population declines and vulnerability to climate change [@Goulson2008-oe; @Soroye2020-zb]. Bumble bees are characteristically cold-adapted species [@Heinrich1994-cv], with peak abundance and diversity in mountain ranges and northern latitudes, where they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Williams1998-xi; @Goulson2010-lo]. Upslope movement of bumble bee species in response to climate change has already been documented, and this is expected to cause the declines or extinction of some bumble bee species [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc]. From a purely geometric perspective, populations of bumble bee species already restricted to the highest elevation zones can be expected to suffer habitat shrinkage [@Dirnbock2011-jf], possibly accompanied by increasing genetic isolation [though see @Fijen2020-zg]. But the outcomes of elevational range shifts on bumble bees also hinge upon the current elevational distributions of bumble bees and their floral resources that will be acted upon by climate change. The upward advance of lowland bumble bee species, for example, may introduce novel competitive pressures on highland species and rewire the pollination networks in which they participate [@Nagamitsu2010-wi; @Brosi2013-mu; @Ishii2013-sm], and elevational [@Lenoir2008-yn] and/or phenological [@Pyke2016-ll] shifts in the floral community could generate mismatches between bumble bee species and their floral hosts. When range shifts are considered in this complex biotic context, it is especially important to understand nonlinearities or thresholds in species-, community-, and interaction-level responses to elevational climate gradients, such as the floristic threshold effect that has been documented at the tree line ecotone [@Descombes2017-id], since these could produce dramatic ecological consequences that would not be predicted by geometric considerations alone.

@@

The peak abundance of all bumble bee species responded significantly to elevation (p ≤ 0.04) **(Figures 3, 4)**. *B. mendax*, *B. monticola*, *B. mucidus*, and *B. pyrenaeus* exhibited a positive response to elevation and peaked in abundance above the tree line (~ 1500 m.s.l.), consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. The abundance of *B. soroeensis* also tended to increase with elevation and peaked above the tree line, but it was also abundant at lower elevations. In contrast, the abundance of *B. hortorum*, *B. pascorum*, and *B. psithyrus* declined with elevation. This effect was especially pronounced in *B. pascuorum*, which was also the most abundant species in our study system overall. *B. gerstaeckeri*, *B. jonellus*, *B. terrestris-lucorum*, and *B. wurflenii* peaked at mid-elevation below the tree line, while *B. pratorum* was unique in exhibiting a sharp peak of abundance directly *at* the tree line.

>>>>

Bumble bees (Hymenoptera: *Bombus* spp.) are classic model organisms in the field of community ecology, but they have more recently become conservation priorities due to well-documented population declines and vulnerability to climate change [@Goulson2008-oe; @Soroye2020-zb]. Bumble bees are characteristically cold-adapted species [@Heinrich1994-cv], with peak abundance and diversity in mountain ranges and northern latitudes, where they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Williams1998-xi; @Goulson2010-lo]. Upslope movement of bumble bee species in response to climate change has already been documented, and this is expected to cause the declines or extinction of some bumble bee species [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; Biella2017-ni; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc]. From a purely geometric perspective, populations of bumble bee species already restricted to the highest elevation zones can be expected to suffer habitat shrinkage [@Dirnbock2011-jf], possibly accompanied by increasing genetic isolation [though see @Fijen2020-zg]. But the outcomes of elevational range shifts on bumble bees also hinge upon the current elevational distributions of bumble bees and their floral resources that will be acted upon by climate change. The upward advance of lowland bumble bee species, for example, may introduce novel competitive pressures on highland species and rewire the pollination networks in which they participate [@Nagamitsu2010-wi; @Brosi2013-mu; @Ishii2013-sm], and elevational [@Lenoir2008-yn] and/or phenological [@Pyke2016-ll] shifts in the floral community could generate mismatches between bumble bee species and their floral hosts. When range shifts are considered in this complex biotic context, it is especially important to understand nonlinearities or thresholds in species-, community-, and interaction-level responses to elevational climate gradients, such as the floristic threshold effect that has been documented at the tree line ecotone [@Descombes2017-id], since these could produce dramatic ecological consequences that would not be predicted by geometric considerations alone.

@@

The peak abundance of all bumble bee species responded significantly to elevation (p ≤ 0.04) **(Figures 3, 4)**. *B. mendax*, *B. monticola*, *B. mucidus*, and *B. pyrenaeus* exhibited a positive response to elevation and peaked in abundance above the tree line (~ 1500 m.s.l.), consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. The abundance of *B. soroeensis* also tended to increase with elevation and peaked above the tree line, but it was also abundant at lower elevations. In contrast, the abundance of *B. hortorum*, *B. pascorum*, and *B. psithyrus* declined with elevation. This effect was especially pronounced in *B. pascuorum*, likely because its exceptionally long nesting cycle is incompatible with the short growing season at high elevations [@Neumayer2009-gx]. *B. gerstaeckeri*, *B. jonellus*, *B. terrestris-lucorum*, and *B. wurflenii* peaked at mid-elevation below the tree line, while *B. pratorum* was unique in exhibiting a sharp peak of abundance directly *at* the tree line.

There’s a nice species-by-species discussion section. That would be a good place to look for descriptive anecdotes to draw upon in our writing. For example: “Bombus pascuorum (SCOPOLI 1763), the field bumblebee, is the most common bumblebee species of the alpine foothills and valley locations, but reaches its altitudinal limit in the area at about 1500 m. It can be assumed that the nest sites are even lower. Because of the extremely long nesting cycle of at least 5 months (REUTER & SCHWAMMBERGER 2002). there are limits to its altitudinal distribution. B. pascuorum was and is found in the ubiquitous distribution in submontane and montane regions and is one of the few bumblebee species which can still be found in cleared agricultural deserts.”... “Bombus pratorum (LINNAEUS 1761), the meadow bumblebee, is a species with an extremely short nest cycle, which can be found from the valley to the summits of the Gesäuse National Park. Its main distribution is in the montane zone, where it is often found in forested and near-forested areas. It is also well documented in the area by historical finds.

### 2021-03-26 ms_2/ms_2_v2.Rmd
almost done ms2...

Bumble bees (Hymenoptera: *Bombus* spp.) have become classic model organisms for the study of competition and species coexistence due to their broad and overlapping distributions and their gross similarity in morphology, behavior, and dietary niche [@Plowright1984-lc; @Goulson2008-zy]. In this role, however, they have proven somewhat enigmatic. On the one hand, there is evidence that resource limitation can be a significant constraint on local bumble communities [@Heinrich1976-wc; @Bowers1985-df], and certain bumble bee communities have been shown to be overdispersed in terms of tongue length or phylogeny, suggesting that competition functions as a community assembly process [@Inouye1977-kl; @Pellissier2013-pk]. Nevertheless, bumble bee communities with an apparent deficit of trait (or phylogenetic) dispersion are commonly observed [@Ranta1981-vx; @Goulson2008-zy]. To account for this inconsistency, @Ranta1981-vx proposed that bumble bees communities are governed by a spatiotemporally heterogeneous mosaic of competitive conditions, such that competitive inequalities between species are sufficiently variable to allow more species to coexist than predicted by trait-based competition models. A complementary explanation, based on the core/satellite species hypothesis [@Hanski1982-hs], is that competitive forces are significant only among regionally abundant and widespread “core species”, and the effects of competition are more pronounced in the composition of regional species pools than in that of local assemblages [@Hanski1982-ie]. In some cases, competition for food may simply be precluded by a saturating abundance of floral resources, implying that populations are limited by other factors, such as abiotic stressors, nest site availability, or predator/pathogen pressure [@Ranta1981-vx; @Williams1989-vr].

@@

While niche overlap cannot serve as a metric of competition *per se* [@Abrams1980-am], it is often used as an index of *potential* competition, particularly in the context of dietary studies [e.g. @French1985-eo; @Steffan-Dewenter2000-ll; @Jung2015-cw], based on the reasoning that if and when food becomes limiting, competition can be expected to vary in proportion to the dietary overlap of co-occurring species [@Pianka1972-oi]. Our study is set in the Berchtesgadener Alps of southeast Germany, a landscape of flower-rich mountain meadows embedded in a matrix of coniferous forest. This system is one of abundant floral resources but harsh abiotic conditions, so it is likely that competition for floral resources is low, or at least transient, and that bumble bee populations are limited instead by the shortness of the growing season. Yet while *realized* competition between bumble bees in our system may be low, the question of *potential* competition is highly relevant because mountain ecosystems are currently subject to at least two powerful drivers of directed change. First, warming climates have already caused measurable upslope shifts in bumble bees and their floral hosts [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc], which could introduce novel competitive interactions between bumble bee species that experience increased sympatry. Second, socioeconomic factors have led to a dramatic decline in traditional alpine grazing management, particularly in the European Alps, resulting in widespread pasture abandonment and reforestation of mountain meadows [@Tappeiner2006-jp; @Tasser2007-qo]. The regression of meadows to forest entails a massive loss of floral resource availability, and even early stages of succession can cause declines in floral abundance and diversity, with corresponding effects on bumble bees [@Walcher2017-kw]. At the same time, more accessible areas at the bases of mountains are undergoing agricultural intensification that displaces natural vegetation and favors grasses over forbs [@Tasser2002-om]. Caught between adverse land use changes above and below and subject to climate-induced range shifts, mountain bumble bees are facing a future of reduced floral resource availability and altered sympatry, and potential competition is likely to become realized according to patterns of niche overlap between species.

@@

Rather than considering various floral morphological traits independently, we relied on the synthesis of @Kugler1970-gb, who developed a typology consisting of ten primary morphotypes and nested subdivision of each. Kugler’s typology was informed by laboratory studies of bumble bee flower choice [@Kugler1943-uy], and there is a precedent for analyzing bumble bee visitation in terms of Kugler’s morphotypes [@Neumayer1999-il; @Schneller2014-ry].

@@

In contrast, visitation of funnel, bell, flag, and head flowers differed significantly both across tongue length classes **(Figure 2)** and between individual species **(Figure 3)**. Funnel flowers were preferred by long-tongued species, though effect was driven mainly by the *Aconitum* specialist *B. gerstaeckeri*. Bell flowers were visited at a higher rate by short-tongued species than by long-tongued species, with medium-tongued species exhibiting intermediate visitation rates. At the species-level, bell flowers were most preferred by the short-tongued species *B. jonellus*, *B. pyrenaeus*, and *B. soroeensis*, and by the medium-tongued species *B. hypnorum* and *B. monticola*. Flag flowers were visited at a higher rate by medium-tongued species than by short-tongued species, with long-tongued species exhibiting intermediate visitation rates. Flag flowers generated strong separation in visitation rate at the species level, receiving very high visitation from the medium-tongued *B. pascuorum* and *B. mucidus*, moderately high visitation from the short-tongued *B. terrestris/lucorum* and *B. lapidarius*, the medium-tongued *B. monticola*, and the long-tongued *B. mendax* and *B. hortorum*, and low visitation by the short-tongued *B. soroeensis*, *B. pyrenaeus*, *B. psithyrus*, *B. pratorum*, and *B. jonellus*, the medium-tongued *B. hypnorum*, and the long-tongued *B. gerstaeckeri*. Models 2 and 3 both explained approximately 31% of total variance (97.5% CI = 30–32%).

```{r fig2, echo = FALSE,  fig.cap = paste("Posterior conditional mean visitation probability parsed by Kugler morphotype and bumble bee tongue length class (Model 2). Point-interval plots depict 66\\% and 95\\% CIs. Letters indicate group differences based on evidence ratio at alpha = 0.05. Note that the range of the x-axis has been rescaled for each panel to emphasize the comparison of visitation probabilities across tongue length classes within floral morphotypes.")}
knitr::include_graphics("../output/brm_pvis_00_plot1_mod.png")
```

```{r fig3, echo = FALSE,  fig.cap = paste("Posterior conditional mean visitation probability by Kugler morphotype and bumble bee species (Model 3). The tongue length class of each species is color-coded. Point-interval plots depict 66\\% and 95\\% CIs. Note that the range of the x-axis has been rescaled for each panel to emphasize the comparison of visitation probabilities across bumble bee species within floral morphptypes.")}
knitr::include_graphics("../output/brm_pvis_02_plot1.png")
```

@@

```{r fig4, echo = FALSE,  fig.cap = paste("Principal coordinates ordination of bumble bee visitation pooled across all samples. To avoid overplotting, only plant genera with an absolute score of at least 0.20 on one or both axes are plotted.")}
knitr::include_graphics("../output/pcoa_aggregate.png")
```

@@

```{r fig5, echo = FALSE,  fig.cap = paste("Niche overlap by level through elevation. Points show empirical niche overlap values, lines show GAM smooths, and shaded bands show 95\% confidence intervals.")}
knitr::include_graphics("../output/niche_gam_00_plot1.png")
```

```{r fig6, echo = FALSE,  fig.cap = paste("Niche overlap by tongue length class difference through elevation. Lines show GAM smooths and shaded bands show 95\% confidence intervals.")}
knitr::include_graphics("../output/niche_gam_20_plot1.png")
```

@@

<!-- Thus, the speculative question of how competition has acted in the historic assembly of bumble bee communities has become an unfolding process observable over ecological time scales and instrumental in the future viability of bumble bee species and their floral mutualists.  -->

<!-- Moreover, the introduction or extinction of species from bumble bee communities, with corresponding addition and release of competitive pressure, can alter the floral visitation patterns of bumble bees and the pollination rates of their host plants [@Brosi2013-mu; @Ishii2013-sm].  -->

<!-- Bumble bees are cold-adapted species [@Heinrich1994-cv] associated mainly with mountain habitats, where they serve as the principal pollinators of many alpine plants [@Williams1998-xi; @Goulson2010-lo]. -->

The question of how bumble bee species with similar tongue lengths coexist hinges on the relationships between species-level floral preferences, trait-matching, and the emergent phenomenon of diet similarity, and this set of relationships can be generalized to describe the niche relations of any group of organisms that share food resources. In this study, we clarify these relationships with respect to bumble bees and wild flowers interacting along a mountain slope elevation gradient. In addition to addressing the question of bumble bee species coexistence, our results suggest how elevational patterns of potential competition may response to climate change, an issue of immediate concern for the conservation of bumble bees and their floral mutualists [@Sirois-Delisle2018-vi; @Soroye2020-zb].

While bumble bees are, indeed, remarkably generalized in their foraging, they are highly discriminating generalists, dramatically over-representing certain floral taxa and morphotypes relative to their abundance while ignoring others. In our study, lip, flag, and head flowers were by far the most strongly preferred when visitation was pooled across all bumble bee species, while bell, disc, and stalk-disc flowers were underrepresented. Patterns of pooled visitation are, of course, driven disproportionately by the behavior of the most abundant bumble bee species, which in our system were *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris/lucorum*, and *B. wurflenii* (Sponsler et al., submitted). Visitation to lip and flag flowers was driven especially by the medium-tongued *B. pascuorum*, the most abundant bumble bee in our study system, and secondarily by the medium-tongued *B. wurflenii*. Visitation to head flowers was driven mainly by the short-tongued *B. soroeensis*.

Tongue length class was an important predictor of preference for some but not all floral morphotypes. For disc and stalk-disc flowers, visitation rates were uniformly low across all tongue length classes. The opposite was true of lip flowers, which were visited at high rates by all tongue length classes. Bell flowers and head flowers were preferred by short-tongued bees, consistent with the shallow corollas of these morphotypes. Conversely, funnel flowers, with their deep corollas were preferred by long-tongued bees, though this effect was driven disproportionately by *B. gerstaeckeri*, which specializes on the funnel flowers of the genus *Aconitum* [@Ponchau2006-af]. Flag flowers were visited at the highest rate by medium-tongued species, and especially by *B. pascuorum* and *B. mucidus*.

@@

@@

@@

As discussed earlier, the interpretation of floral visitation rates and diet overlap hinge strongly on the extent to which interspecific competition actually occurs in our study system. We have argued that the assumption of minimal competition is reasonable in the context of the Berchtesgadener Alps, owing to their rich floral resources and restrictive abiotic conditions, together with the manifest co-occurrence of bumble bee species with similar tongue lengths. This leads to the interpretation of abundance-conditioned visitation rates as *preferences* and the interpretation of diet overlap as *potential competition*. To the extent that interspecific competition *does* occur in our system, though, visitation rates would represent floral preference distorted by the pressure of competition toward niche partitioning, rendering diet overlap an underestimate of potential competition. While we maintain, based on our knowledge of our study system, that the first interpretation is more likely, both interpretations are plausible and must be kept in mind. Given our assumption of negligible interspecific competition, however, it might be questioned whether the inference of *potential* competition is ecologically relevant, since only realized competition can cause competitive exclusion. In answer to this question, we point out that while the assumption of negligible interspecific competition is necessary for the inferences we make concerning floral preference and potential competition, the processes of (1) imperfect trait-matching bewteen bumble bee tongue length and floral morphotype and (2) taxonomic subpartitioning of floral morphotypes that we infer in our analysis still obtain under conditions of realized interspecific competition, even if their inference under those conditions would be obscured by the confounding effects of competitive niche partitioning. Indeed, conditions of realized competition likely occur in all bumble bee communities, including those of our study system, but these conditions may be spatially and/or temporally localized [@Ranta1981-vx], and perhaps associated with extreme climatic events that cause dramatic reductions in floral resource avaulability [@Rasmont2012-zq; @Thomson2016-gq; @Phillips2018-me]. This calls to mind the insight of MacArthur1972-qz, that competition need not be frequent to have lasting ecological impacts. Potential competition in the form of dietary overlap is perhaps the best indication of the effects that future instances of realized competition will have, and it is likely that the "ghosts of competition past" [@Connell1980-dn] haunt the floral preferences exhibited by bumble bees today.

Empirical measurements of niche overlap are, however, inherently ambiguous with respect to competition.

Focusing on *potential* competition does not, however, escape the equivocity of niche overlap, since an observed pattern of niche overlap could just as plausibly be interpreted as adaptive niche partitioning in response to *realized* competition [@Pianka1974-qk]. This creates a predicament in which niche overlap can only be interpreted as an index of potential competition when realized competition is assumed to be negligible, an assumption that can only be justified on the basis of additional knowledge of the study system under consideration.

Our study is set in the Berchtesgadener Alps of southeast Germany, a landscape of flower-rich mountain meadows embedded in a matrix of coniferous forest. This system is one of rich floral resources but harsh abiotic conditions, so it is likely that competition for floral resources is low, or at least transient, and that bumble bees are limited instead by the shortness of the growing season and by weather fluctuations during the critical period of spring colony founding. On this basis, we argue for the interpretation of niche overlap as an index of *potential* competition rather than a response to *realized* competition, with the caveat this interpretation may misrepresent certain temporal intervals (e.g. early spring) or elevational bands (e.g. the sparse alpine grasslands above the tree line ecotone) [Sponsler et al., submitted].

# Notes from conversation with Kalli

- Explain overall preferences in figure 1 by noting that even short-tongued bumble bees represent the long-end of the complete pollinator spectrum; deeper flowers tend to have higher nectar rewards, and open flowers may be subject to competition from other visitors like Syrphids.

- Tying competition back into the discussion. The premise of the paper is that we are modeling *potential competition* with the idea that it predicts realized competition if and when conditions change and competition ensues. So far, the main thrust has been climate change and how it could induce competition via changes in species distributions. But that is probably not the most important driver in Berchtesgaden. Prior to human influence, the landscape was mostly forested, and floral resources were presumably scarce, perhaps concentrated in forest gaps created by fire or tree falls or avalanches or whatever. When humans cleared the land for grazing, they created the flower-rich meadow habitats that now characterize the landscape. More recently, though, economic drivers have caused a withdraw of human management from mountain meadows and, consequently, reforestation [@Tasser2007-qo]. At the same time, land use has intensified at lower elevations. So, mountain bumble bees may find themselves caught between the forest and the farm, so to speak, experience loss of habitat at high elevations due to the withdraw of human activity and loss of habitat at low elevations due to the intensification of human management. We can be confident that if these trends persist, competition will come into play. There is a wealth of literature on pasture abandonment and reforestation in the Alps, including at least one paper that deals explicitly with bumble bees (lots more that deal with floral diversity).

- Explain that traits such as color (really especially color) probably explain much of the variance in preference not explained by morphology.

>>>>

Bumble bees (Hymenoptera: *Bombus* spp.) have become classic model organisms for the study of competition and species coexistence due to their broad and overlapping distributions and their gross similarity in morphology, behavior, and dietary niche [@Plowright1984-lc; @Goulson2008-zy]. In this role, however, they have proven somewhat enigmatic. On the one hand, there is evidence that resource limitation can be a significant constraint on local bumble communities [@Heinrich1976-wc; @Bowers1985-df], and certain bumble bee communities have been shown to be overdispersed in terms of tongue length or phylogeny, suggesting that competition functions as a community assembly process [@Inouye1977-kl; @Pellissier2013-pk]. Nevertheless, bumble bee communities with an apparent deficit of trait (or phylogenetic) dispersion are commonly observed [@Ranta1981-vx; @Goulson2008-zy; Iserbyt2008-jn]. To account for this inconsistency, @Ranta1981-vx proposed that bumble bees communities are governed by a spatiotemporally heterogeneous mosaic of competitive conditions, such that competitive inequalities between species are sufficiently variable to allow more species to coexist than predicted by trait-based competition models. A complementary explanation, based on the core/satellite species hypothesis [@Hanski1982-hs], is that competitive forces are significant only among regionally abundant and widespread “core species”, and the effects of competition are more pronounced in the composition of regional species pools than in that of local assemblages [@Hanski1982-ie]. In some cases, competition for food may simply be precluded by a saturating abundance of floral resources, implying that populations are limited by other factors, such as abiotic stressors, nest site availability, or predator/pathogen pressure [@Ranta1981-vx; @Williams1989-vr; Iserbyt2008-jn].

@@

While niche overlap cannot serve as a metric of competition *per se* [@Abrams1980-am], it is often used as an index of *potential* competition, particularly in the context of dietary studies [e.g. @French1985-eo; @Steffan-Dewenter2000-ll; @Jung2015-cw], based on the reasoning that if and when food becomes limiting, competition can be expected to vary in proportion to the dietary overlap of co-occurring species [@Pianka1972-oi]. Our study is set in the Berchtesgadener Alps of southeast Germany, a landscape of flower-rich mountain meadows embedded in a matrix of coniferous forest. This system is one of abundant floral resources but harsh abiotic conditions, so it is likely that competition for floral resources is low, or at least transient, and that bumble bee populations are limited instead by the shortness of the growing season, as has been suggested in the similar environment of the Eyne Valley of France [@Iserbyt2008-jn]. Yet while *realized* competition between bumble bees in our system may be low, the question of *potential* competition is highly relevant because mountain ecosystems are currently subject to at least two powerful drivers of directed change. First, warming climates have already caused measurable upslope shifts in bumble bees and their floral hosts [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc], which could introduce novel competitive interactions between bumble bee species that experience increased sympatry. Second, socioeconomic factors have led to a dramatic decline in traditional alpine grazing management, particularly in the European Alps, resulting in widespread pasture abandonment and reforestation of mountain meadows [@Tappeiner2006-jp; @Tasser2007-qo]. The regression of meadows to forest entails a massive loss of floral resource availability, and even early stages of succession can cause declines in floral abundance and diversity, with corresponding effects on bumble bees [@Iserbyt2015-jc; @Walcher2017-kw; @Suzuki-Ohno2020-jv]. At the same time, more accessible areas at the bases of mountains are undergoing agricultural intensification that displaces natural vegetation and favors grasses over forbs [@Tasser2002-om]. Caught between adverse land use changes above and below and subject to climate-induced range shifts, mountain bumble bees are facing a future of reduced floral resource availability and altered sympatry, and potential competition is likely to become realized according to patterns of niche overlap between species.

@@

Rather than considering various floral morphological traits independently, we relied on the synthesis of @Kugler1970-gb, who developed a typology consisting of ten primary morphotypes and nested subdivision of each. Kugler’s typology was informed by laboratory studies of bumble bee flower choice [@Kugler1943-uy], and there is a precedent for analyzing bumble bee visitation in terms of Kugler’s morphotypes [@Farnberger1996-xo; @Neumayer1999-il; @Schneller2014-ry].

@@

In contrast, visitation of funnel, bell, flag, and head flowers differed significantly both across tongue length classes **(Figure 2)** and between individual species **(Figure 3)**. Funnel flowers were preferred by long-tongued species, though effect was driven mainly by the *Aconitum* specialist *B. gerstaeckeri* [@Ponchau2006-af; @Schneller2014-ry]. Bell flowers were visited at a higher rate by short-tongued species than by long-tongued species, with medium-tongued species exhibiting intermediate visitation rates. At the species-level, bell flowers were most preferred by the short-tongued species *B. jonellus*, *B. pyrenaeus*, and *B. soroeensis*, and by the medium-tongued species *B. hypnorum* and *B. monticola*. Flag flowers were visited at a higher rate by medium-tongued species than by short-tongued species, with long-tongued species exhibiting intermediate visitation rates. Flag flowers generated strong separation in visitation rate at the species level, receiving very high visitation from the medium-tongued *B. pascuorum* and *B. mucidus*, moderately high visitation from the short-tongued *B. terrestris/lucorum* and *B. lapidarius*, the medium-tongued *B. monticola*, and the long-tongued *B. mendax* and *B. hortorum*, and low visitation by the short-tongued *B. soroeensis*, *B. pyrenaeus*, *B. psithyrus*, *B. pratorum*, and *B. jonellus*, the medium-tongued *B. hypnorum*, and the long-tongued *B. gerstaeckeri*. Models 2 and 3 both explained approximately 31% of total variance (97.5% CI = 30–32%).

```{r fig2, echo = FALSE,  fig.cap = paste("Posterior conditional mean visitation probability parsed by Kugler morphotype and bumble bee tongue length class (Model 2). Point-interval plots depict 66\\% and 95\\% CIs. Letters indicate group differences based on evidence ratio at alpha = 0.05. Note that the range of the x-axis has been rescaled for each panel to emphasize the comparison of visitation probabilities across tongue length classes within floral morphotypes.")}
knitr::include_graphics("../output/brm_pvis00_plot1_mod2.png")
```

```{r fig3, echo = FALSE,  fig.cap = paste("Posterior conditional mean visitation probability by Kugler morphotype and bumble bee species (Model 3). The tongue length class of each species is color-coded. Point-interval plots depict 66\\% and 95\\% CIs. Note that the range of the x-axis has been rescaled for each panel to emphasize the comparison of visitation probabilities across bumble bee species within floral morphptypes.")}
knitr::include_graphics("../output/brm_pvis_02_plot1_mod.png")
```

@@

```{r fig4, echo = FALSE,  fig.cap = paste("Principal coordinates ordination of bumble bee visitation pooled across all samples. To avoid overplotting, only plant genera with an absolute score of at least 0.20 on one or both axes are plotted.")}
knitr::include_graphics("../output/pcoa_aggregate_mod.png")
```

@@

```{r fig5, echo = FALSE,  fig.cap = paste("Niche overlap by level through elevation. Points show empirical niche overlap values, lines show GAM smooths, and shaded bands show 95\\% confidence intervals.")}
knitr::include_graphics("../output/niche_gam_00_plot1.png")
```

```{r fig6, echo = FALSE,  fig.cap = paste("Niche overlap by tongue length class difference through elevation. Lines show GAM smooths and shaded bands show 95\\% confidence intervals.")}
knitr::include_graphics("../output/niche_gam_20_plot1.png")
```

@@

Any discussion of competition and niche partitioning among *Bombus* species must begin with the observation that bumble bees are unique among insect pollinators in the degree to which their foraging combines contrasting extremes of bias and flexibility. Coexistence between bumble bee species with similar tongue lengths hinges on the relationships between species-level floral preferences, trait-matching, and the emergent phenomenon of diet similarity. In this study, we clarify these relationships with respect to bumble bees and wild flowers interacting along a mountain slope elevation gradient. In addition to addressing the question of bumble bee species coexistence, our results suggest how patterns of competition may respond to climate and land use change along mountain slopes, an issue of immediate concern for the conservation of bumble bees and their floral mutualists [@Sirois-Delisle2018-vi; @Soroye2020-zb].

## Floral preference is only partly explained by the relationship between bumble bee tongue length and floral morphotype

 While biases can be explained partly in terms of morphological matching between tongue length and floral morphotype

 The rule is flexibility, and floral morphotypes do not constitute hard boundaries to bumble bee flower choice. Nevertheless, we demonstrate that for certain floral morphotypes --- funnel flower, bell flowers, flag flowers, and head flower --- visitation rate is affected strongly by tongue length.

In our study, lip, flag, and head flowers were by far the most strongly preferred when visitation was pooled across all bumble bee species, while disc, funnel, bell, and stalk-disc flowers were comparatively underrepresented. These findings are broadly consistent with the aggregate visitation patterns previously reported by @Schneller2014-ry in the central Alps at Vorarlberg and Neumayer1999-il ~50 km south of our study system, suggesting that our findings are at least regionally generalizeable. When considering these patterns from the perspective of pollination services to mountain flora, however, it is important to note that while bumble bees are the dominant Hymenopteran pollinators in our study system, Syrphid flies are also abundance flower visitors whose visitation is mainly restricted to the open flower morphotypes (e.g. disc flowers) that receive less attention from bumble bees [@Neumayer1999-il]. In the context of our research question, we have emphasized the differences in tongue length between bumble bee species, but all bumble bees have relatively long tongues compared to the overall pollinator community, and the process of niche partitioning *among* bumble bee species is nested within the broader process of niche partitioning *between* bumble bees and other flower visitors.

@@

<!-- - Explain that traits such as color (really especially color) probably explain much of the variance in preference not explained by morphology [@Neumayer1999-il].  -->

## The influence of tongue length on floral preference does not carry on to genus-level niche overlap, and floral diversity within morphotypes relaxes potential competition between bumble bee species of similar tongue length

@@

<!-- @Durieux2000-uk similarly ofund that while tongue length is correlated with floral preference, it is not correlated with diet similarity. But we did better stats! -->

## Climate, land use, and the future of bumble bee niche relations

@@

<!-- As discussed earlier, the interpretation of floral visitation rates and diet overlap hinge strongly on the extent to which interspecific competition actually occurs in our study system. We have argued that the assumption of minimal competition is reasonable in the context of the Berchtesgadener Alps, owing to their rich floral resources and restrictive abiotic conditions, together with the manifest co-occurrence of bumble bee species with similar tongue lengths. This leads to the interpretation of abundance-conditioned visitation rates as *preferences* and the interpretation of diet overlap as *potential competition*. To the extent that interspecific competition *does* occur in our system, though, visitation rates would represent floral preference distorted by the pressure of competition toward niche partitioning, rendering diet overlap an underestimate of potential competition. While we maintain, based on our knowledge of our study system, that the first interpretation is more likely, both interpretations are plausible and must be kept in mind. Given our assumption of negligible interspecific competition, however, it might be questioned whether the inference of *potential* competition is ecologically relevant, since only realized competition can cause competitive exclusion. In answer to this question, we point out that while the assumption of negligible interspecific competition is necessary for the inferences we make concerning floral preference and potential competition, the processes of (1) imperfect trait-matching bewteen bumble bee tongue length and floral morphotype and (2) taxonomic subpartitioning of floral morphotypes that we infer in our analysis still obtain under conditions of realized interspecific competition, even if their inference under those conditions would be obscured by the confounding effects of competitive niche partitioning. Indeed, conditions of realized competition likely occur in all bumble bee communities, including those of our study system, but these conditions may be spatially and/or temporally localized [@Ranta1981-vx], and perhaps associated with extreme climatic events that cause dramatic reductions in floral resource avaulability [@Rasmont2012-zq; @Thomson2016-gq; @Phillips2018-me]. This calls to mind the insight of MacArthur1972-qz, that competition need not be frequent to have lasting ecological impacts. Potential competition in the form of dietary overlap is perhaps the best indication of the effects that future instances of realized competition will have, and it is likely that the "ghosts of competition past" [@Connell1980-dn] haunt the floral preferences exhibited by bumble bees today. -->

<!-- Empirical measurements of niche overlap are, however, inherently ambiguous with respect to competition.  -->

<!-- Focusing on *potential* competition does not, however, escape the equivocity of niche overlap, since an observed pattern of niche overlap could just as plausibly be interpreted as adaptive niche partitioning in response to *realized* competition [@Pianka1974-qk]. This creates a predicament in which niche overlap can only be interpreted as an index of potential competition when realized competition is assumed to be negligible, an assumption that can only be justified on the basis of additional knowledge of the study system under consideration.  -->

<!-- - Explain that traits such as color (really especially color) probably explain much of the variance in preference not explained by morphology.  -->

<!-- Bumble bees are cold-adapted species [@Heinrich1994-cv] associated mainly with mountain habitats, where they serve as the principal pollinators of many alpine plants [@Williams1998-xi; @Goulson2010-lo]. The introduction or extinction of species from bumble bee communities, with corresponding addition and release of competitive pressure, can alter the floral visitation patterns of bumble bees and the pollination rates of their host plants [@Brosi2013-mu; @Ishii2013-sm].  -->

<!-- Lip and head flowers have the highest nectar reward per inflorescence [@Neumayer1999-il]. -->
<!-- Interestingly, @Farnberger1996-xo in a localized study near Wien found rather different results. Lip flowers were still found to be the favorite overall, and head flowers were popular, too, but flag flowers were of secondary importance, and bell flowers were the most visited morphotype for humi, pasc, and prat. Pasc almost totally ignored bell flowers in our study. -->
